---
title: 'Prediction of keyboard input: Exploratory data analysis and development plan'
author: "Daigo Tanaka"
date: "March 21, 2015"
output: html_document
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# My usual front matter for Rmd

setwd("/Users/daigo/projects/stats/datascience-capstone")
message(paste("Working directory:", getwd(), sep=" "))

library(knitr)
library(RCurl)
library(stringr)
library(ggplot2)
library(gridExtra)

version <- sessionInfo()$R.version$version.string
platform <- sessionInfo()$platform

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, comment=NA,
              results="asis", tidy=FALSE, cache=FALSE)

# Set significant digits
options(scipen = 20, digits = 2)

# Load caption helper
code <- getURL("https://gist.githubusercontent.com/daigotanaka/17930c2ff891e05a83f5/raw/3c4f4af54cbdf8ec77d4b5112b34e9307ce92b7c/rmd_caption_helper.R")
eval(parse(text=code))

fig <- Caption$new()
tab <- Caption$new("Table")
fn <- Footnote$new()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Helper scripts

# Clean up and tokenize
source("cleanup_tokenize.R", echo=FALSE)

# File name helpers
source("filename.R", echo=FALSE)

# Create train and test data
source("create_train_test.R", echo=FALSE)

# Create n-grams
source("create_ngrams.R", echo=FALSE)

# prediction tools
source("prediction.R", echo=FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Download data
source("download_data.R", echo=FALSE)

locale <- "en_US"
source <- c("news", "blogs", "twitter")

# Generate training & test data
set.seed(926)
for (i in 1:length(source)) {
    trainFileName <- paste(dataDir, "/", locale, "/", locale, ".", source[i],
                           "_train.txt", sep="")
    if (file.exists(trainFileName)) {
        message(paste("File already exists for ", locale, "_", source[i],
                      sep=""))
        next
    }
    testFileName <- paste(dataDir, "/", locale, "/", locale, ".", source[i],
                           "_test.txt", sep="")
    inFileName <- getFileName(locale, source[i])
    createTrainTestData(inFileName, trainFileName, testFileName, cleanTrain=TRUE)
    message(paste("Generated ", source[i], " train/test data"), sep="")
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load or generate n-grams from the training data
locale <- "en_US"
trainSource <- c("news_train", "blogs_train", "twitter_train")
fourGramChunksFileName <- paste("models/fourGramChunks_",
                                locale,
                                "_",
                                paste(trainSource, collapse="_"),
                                ".rds",
                                sep="")

if (!file.exists("models/fourGrams_en_US_news_blogs_twitter.rds")){
    if (file.exists(fourGramChunksFileName)) {
        message("Loading four-gram chunks from file...")
        system.time(fourGramChunks <- as.data.table(readRDS(fourGramChunksFileName)))
    } else {
        message("Generating four-gram chunks...")
        dictFile <- "models/dictionary.rds"
        if (file.exists(dictFile)) {
            dict <- readRDS(dictFile)
            dictenv <- new.env(hash=TRUE, size=length(dict))
            lapply(rep(1:length(dict)),
                   FUN=function(i) {assign(dict[i], i, envir=dictenv)})
            rm(dict)
            gc()
        }
        system.time(fourGramChunks <- create4GramChunks(locale=locale, source=trainSource))
        gc()
        system.time(
            saveRDS(
                fourGramChunks,
                fourGramChunksFileName))
        gc()
    }

    message("Generating n-gram models...")
    system.time(fourGrams <- fourGramChunks[!is.na(N2) & !is.na(N3) & !is.na(N4), sum(N), .(N1, N2, N3, N4)])
    fourGrams <- fourGrams[order(-fourGrams$V1),]
    saveRDS(fourGrams, "models/fourGrams_en_US_news_blogs_twitter.rds")
    rm(fourGrams)
    gc()

    system.time(
        threeGrams <- fourGramChunks[!is.na(N2) & !is.na(N3),  sum(N),
                                     .(N1, N2, N3)])
    threeGrams <- threeGrams[order(-threeGrams$V1),]
    saveRDS(threeGrams, "models/threeGrams_en_US_news_blogs_twitter.rds")
    rm(threeGrams)
    gc()

    system.time(
        twoGrams <- fourGramChunks[!is.na(N2),  sum(N), .(N1, N2)])
    twoGrams <- twoGrams[order(-twoGrams$V1),]
    saveRDS(twoGrams, "models/twoGrams_en_US_news_blogs_twitter.rds")
    rm(twoGrams)
    gc()

    system.time(
        oneGrams <- fourGramChunks[,  sum(N), .(N1)])
    oneGrams <- oneGrams[order(-oneGrams$V1),]
    saveRDS(oneGrams, "models/oneGrams_en_US_news_blogs_twitter.rds")
    rm(oneGrams)
    rm(fourGramChunks)
    gc()
}
# gc()
message("Loading n-gram models from disk...")
nGrams <- list(readRDS("models/oneGrams_en_US_news_blogs_twitter.rds"),
               readRDS("models/twoGrams_en_US_news_blogs_twitter.rds"),
               readRDS("models/threeGrams_en_US_news_blogs_twitter.rds"),
               readRDS("models/fourGrams_en_US_news_blogs_twitter.rds"))

if (!file.exists("models/dictionary.rds")) {
    message("Generating dictionary file...")
    dict <- as.vector(nGrams[[1]]$N1)
    saveRDS(dict, "models/dictionary.rds")
    rm(dict)
    gc()
}
message("Loading dictionary file...")
dict <- readRDS("models/dictionary.rds")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Count lines and words
counts <- data.frame(
    locale=c(rep("en_US", 6), rep("de_DE", 6), rep("ru_RU", 6), rep("fi_FI", 6)),
    source=rep(c("news", "blogs", "twitter"), 8),
    type=rep(c("lines", "words"), 12),
    count=rep(0, 24)
    )

getCount <- function(locale, source, type) {
    return (counts[
                counts$locale==locale &
                counts$source==source &
                counts$type==type,]$count)
}

for (i in 1:nrow(counts)) {
    item <- counts[i,]
    inFileName <- getFileName(item$locale, item$source)

    if (item$type == "words") {
        ret <- system(paste("wc -w ", getwd(), "/", inFileName, sep=""),
                             intern=TRUE)
        counts[i,]$count <- as.integer(strsplit(str_trim(ret), " ")[[1]][1])
    } else {
        ret <- system(paste("wc -l ", getwd(), "/", inFileName, sep=""),
                         intern=TRUE)
        counts[i,]$count <- as.integer(strsplit(str_trim(ret), " ")[[1]][1])
    }
}
rm(item)
# gc()

# Most frequent n-grams
ngrams <- c()
for (i in 1:4) {
    for (j in 1:5) {
        ngrams <- c(ngrams, paste(nGrams[[i]][j], collapse=" "))
    }
}
mf <- matrix(ngrams, nrow=4, ncol=5, byrow=TRUE)

# > object.size(nGrams[[4]])
# 1282268544 bytes
```
<h1 class="countheads">Prediction of keyboard input: Exploratory data analysis and development plan</h1>

***Daigo Tanaka***

***March 21, 2015***

## Introduction

To assist typing texts in small devices such as smart phones, a software
keyboard could predict what the user is going to type and display the candidates
of words so that the user can choose instead of typing the whole word.
Statistical machine learning system can perform such prediction tasks by
analyzing example texts (corpus). This report
```{r}
fn$label("source")
```
describes a plan for developing
keyboard input prediction algorithm.

The current report is a milestone report as part to complete Data Science
Specialization capstone project on Coursera instructed by Jeff Leek, PhD, Roger D.
Peng, PhD, and Brian Caffo, PhD from Department of Biostatistics at Johns Hopkins
Bloomberg School of Public Health
```{r}
fn$label("coursera")
```
in collaboration with SwiftKey
```{r}
fn$label("swiftkey")
```
.

## Data aquisition and transformation

The corpus data
```{r}
fn$label("dataset")
```
used in this report is generated with a freely available data
from HC corpora project
```{r}
fn$label("hccorpora")
```
. The corpora are collected from publicly available sources by a web crawler.
The sources include news, blogs, and Twitter. The locale includes US English
(en_US), German (de_DE), Russian (ru_RU), and Finnish (fi_FI). Table
`r tab$label("counts")`
shows the line and word counts from each locale and source. In this milestone
report, I focus on analyzing en_US data.

`r render_caption(tab$text("counts","Line and word counts from each locale and source"))`

| Locale | Source  | # of lines | # of words |
| :----- | :------ | ---------: | ---------: |
| en_US  | news    | `r getCount("en_US", "news", "lines")` | `r getCount("en_US", "news", "words")`|
|        | blogs    | `r getCount("en_US", "blogs", "lines")` | `r getCount("en_US", "blogs", "words")`|
|        | twitter | `r getCount("en_US", "twitter", "lines")` | `r getCount("en_US", "twitter", "words")`|
| de_DE  | news    | `r getCount("de_DE", "news", "lines")` | `r getCount("de_DE", "news", "words")`|
|        | blogs    | `r getCount("de_DE", "blogs", "lines")` | `r getCount("de_DE", "blogs", "words")`|
|        | twitter | `r getCount("de_DE", "twitter", "lines")` | `r getCount("de_DE", "twitter", "words")`|
| ru_RU  | news    | `r getCount("ru_RU", "news", "lines")` | `r getCount("ru_RU", "news", "words")`|
|        | blogs    | `r getCount("ru_RU", "blogs", "lines")` | `r getCount("ru_RU", "blogs", "words")`|
|        | twitter | `r getCount("ru_RU", "twitter", "lines")` | `r getCount("ru_RU", "twitter", "words")`|
| fi_FI  | news    | `r getCount("fi_FI", "news", "lines")` | `r getCount("fi_FI", "news", "words")`|
|        | blogs    | `r getCount("fi_FI", "blogs", "lines")` | `r getCount("fi_FI", "blogs", "words")`|
|        | twitter | `r getCount("fi_FI", "twitter", "lines")` | `r getCount("fi_FI", "twitter", "words")`|

### Generating training and test data

For each data source, 60% of the input was randomly picked to form sets of
training data, and the rest is stored as test data. The rest of the report
only analyze the training data.

### Cleaning data

The following words are considered during the data cleaning process:

- **Non-ASCII characters**: There is not an occasion for human to type non-ASCII
  characters, so they are removed from the training data.
- **Profane words**: While the distribution of profane words may be statistically
  interesting, I would like to avoid suggesting such words in the keyboard input
  application as it may be offensive to many users. A list of profane words in
  English was obtained from Luis von Ahn's research group
```{r}
fn$label("profane")
```
  . The profane words
  in the training data was replaced by "****" during the training data generation.
- **Stop words**: When extracting the semantics of the text for applications
  such as document classification, it is useful to remove stop words entirely.
  In the keyboard input prediction however, it may be helpful to suggest stop
  words (e.g. Suggesting "herself" followed by the existing input "She did this
  all by") So, I am currently creating training data with and without stop
  words included. The rest of the report shows the result with stop words
  included. The list of stop words were obtained from XPO6's list
```{r}
fn$label("stopwords")
```
  .
- **Stemming**: I chose not to stem the words for similar reason as in stop words.
  There are common expressions (e.g. killing two birds with one stone), where
  singularization won't work well.

### Tokenizing and generating 4-gram chunks

The input often contains multiple sentences and fragments that are typically
separated by characters such as ".", "?", "!", ",", ";", and ":". The input was
made into sentence/fragment chunks with those separation characters. One of the
unresolved problems of course is the abbreviations such as "U.S.A." and "Ph.D.".

The sentence/fragment chunks are then tokenized. For keyboard input prediction
purpose, the apostrophes "'" are treated as a valid characters to preserve the
expression like "don't" and "can't" as these are common expressions that the
users could be benefit by automated suggestions instead of having to type "'".

The goal in the current report was to create 4, tri, bi, and uni-grams
efficiently. There are R packages to create n-grams, but those tend to be
over-sophisticated for the current application, and often found to be very slow
in execution and consuming large memory. So, this task was accomplished by
writing my own function. The function combines all the tokens into a vector
object, inserting 3 NAs at each break of sentences/fragments. By copying the
vector into a data table with four token columns by shifting one word as it
moves onto the next row, 4 gram chunks can be created efficiently. Here are
some examples of the chunks:

1. He started playing guitar
2. started playing guitar NA
3. playing guitar NA NA
4. guitar NA NA NA

In the example above, the first token from all are used to generate unigrams.
The sequence of the first and second tokens from Example 1 to 3 are used to
to generate bigrams. The sequence of the first, second, and third from Example
1 and 2 are used to generated trigrams. The whole sequence from the Example 1
is used to generate 4-grams. Finally, the data table is aggregated by counting
the occurrences of each n-grams.

## Exploratory data analysis

### Most frequent n-grams

Table `r tab$label("mostfreq")` shows the 5
most frequent n-grams. The frequent appearance of the expression such as "thanks
for the follow" indicates that Twitter corpus is biasing the distribution.

`r render_caption(tab$text("mostfreq", "Five most frequent word sequences in 1 to 4-grams. The number indicates the frequency in the training data."))`

| n-gram  | No.1         | No.2         | No.3         | No.4         | No.5         |
| :------ | :----------- | :----------- | :----------- | :----------- | :----------- |
| Unigram | `r mf[1, 1]` | `r mf[1, 2]` | `r mf[1, 3]` | `r mf[1, 4]` | `r mf[1, 5]` |
| Bigram  | `r mf[2, 1]` | `r mf[2, 2]` | `r mf[2, 3]` | `r mf[2, 4]` | `r mf[2, 5]` |
| Trigram | `r mf[3, 1]` | `r mf[3, 2]` | `r mf[3, 3]` | `r mf[3, 4]` | `r mf[3, 5]` |
| 4-gram  | `r mf[4, 1]` | `r mf[4, 2]` | `r mf[4, 3]` | `r mf[4, 4]` | `r mf[4, 5]` |

### Distribution of n-grams

Figure
`r fig$label("distributions")`
shows the distributions of n-gram frequency. They all look like an exponential distribution. In fact, the majority of the 4-gram and trigram entities appear only once as in
Table `r tab$label("once")`. This is a useful feature of the data when considering the
strategy for efficient modeling as discussed in the later section.

```{r, html.cap=fig$text("distributions", "The distribution of n-gram frequency. The count is in log(10) scale.")}
# n-gram distribution
grid.arrange(
    ggplot(data=data.frame(frequency=nGrams[[1]]$V1), aes(x=frequency)) +
        geom_histogram(drop=TRUE) + scale_y_log10() +
        ggtitle(paste(1, "-gram")),
    ggplot(data=data.frame(frequency=nGrams[[2]]$V1), aes(x=frequency)) +
        geom_histogram(drop=TRUE) + scale_y_log10() +
        ggtitle(paste(2, "-gram")),
    ggplot(data=data.frame(frequency=nGrams[[3]]$V1), aes(x=frequency)) +
        geom_histogram(drop=TRUE) + scale_y_log10() +
        ggtitle(paste(3, "-gram")),
    ggplot(data=data.frame(frequency=nGrams[[4]]$V1), aes(x=frequency)) +
        geom_histogram(drop=TRUE) + scale_y_log10() +
        ggtitle(paste(4, "-gram"))
    , ncol=2)
```

`r render_caption(tab$text("once", "In bi, tri, and 4-grams, the majority of the entities appear only once."))`

| n-grams | % of n-grams appearing only once |
| :------ | -------------------------------: |
| Unigram | `r 100 * nGrams[[1]][V1==1, sum(V1)] / nGrams[[1]][, sum(V1)]` |
| Bigram  | `r 100 * nGrams[[2]][V1==1, sum(V1)] / nGrams[[2]][, sum(V1)]` |
| Trigram | `r 100 * nGrams[[3]][V1==1, sum(V1)] / nGrams[[3]][, sum(V1)]` |
| 4-gram  | `r 100 * nGrams[[4]][V1==1, sum(V1)] / nGrams[[4]][, sum(V1)]` |

## Plan for developing prediction algorithm

### Language model generation

Generally, the text prediction problem can be expressed as the calculation of
the conditional probability of a n-gram that are approximated by the frequency
counts of the training data:

$$P(w_{i}|w_{i-(n -1)},...,w_{i-1}) = \frac{count(w_{i-(n -1)},...,w_{i-1}, w_{i})}{count(w_{i-(n -1)},...,w_{i-1})}$$

The prediction should be chosen from the n-th word in the n-gram with the
highest probability. In case we did not encounter the particular n-gram in the
training data, so called Stupid Backoff
```{r}
fn$label("stupidbackoff")
```
is used to approximate the probability. In Stupid Backoff, the conditional
probability is recursively approximated as:

$$S(w_{i}|w_{i-(n-1)},...,w_{i-1}) = \frac{count(w_{i-(n-1)},...,w_{i-1}, w_{i})}{count(w_{i-(n-1)},...,w_{i-1})}\ if\ count(w_{i-(n-1)},...,w_{i-1}, w_{i}) > 0,$$
$$\alpha S(w_{i}|w_{i-(n-2)},...,w_{i-1})\ otherwise$$

, where $\alpha$ is an empirically chosen constant typically set to 0.4.

### Efficient modeling strategy

An insight I got from the exploratory data analysis is that the most bi, tri,
and 4-grams appear only once, making the majority of the rows of the n-gram
data table filled with count 1. By using techniques such as Kneser–Ney
smoothing
```{r}
fn$label("knsmooth")
```
, the probability of the low frequency n-grams is rather derived from
the continuous probability from the lower order n-grams. This will allow the
data table to drop all the entries with single count. Such smoothing techniques
also enable the calculation of the n-grams that are not included in the
training data.

```{r}
fn$update("source", 'The R markdown source code generated this document is availalbe on <a href="https://github.com/daigotanaka/datascience-capstone/blob/master/keyboard-prediction.Rmd" target="_blank">github repository</a>.')
fn$update("coursera", 'Coursera Data Science Specialization Capstone project [<a href="https://www.coursera.org/course/dsscapstone">link</a>]')
fn$update("swiftkey", 'Swiftkey [<a href="http://swiftkey.com/">link</a>]')
fn$update("dataset", 'Dataset can be downloaded from <a href="https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip">here</a>. (zipped, 548MB)')
fn$update("hccorpora", 'HC corpora [<a href="http://www.corpora.heliohost.org/">link</a>]')
fn$update("profane", 'Offensive/Profane Word List [<a href="http://www.cs.cmu.edu/~biglou/resources/">link</a>]')
fn$update("stopwords", 'List of English Stop Words - XPO6 [<a href="http://xpo6.com/list-of-english-stop-words/">link</a>]')
fn$update("stupidbackoff", 'Brants et al, Large language models in machine translation, EMNLP (2007), pp 858--867 [<a href="http://www.aclweb.org/anthology/D07-1090.pdf">link</a>]')
fn$update("knsmooth", 'Chen, et al. An empirical study of smoothing techniques for language modeling. Computer Speech and Language (1999) 13, pp.359-394. [<a href="http://u.cs.biu.ac.il/~yogo/courses/mt2013/papers/chen-goodman-99.pdf">link</a>]')

fn$render(head='<h3 class="nocount">References and notes</h3>')
```
